---
title: "Covid_19 Trends"
author: "Mbong Mou Carine"
date: "`r Sys.Date()`"
output: html_document
---

## Investigating COVID-19 Virus Trends

## Introduction

The COVID-19 pandemic, caused by the SARS-CoV-2 virus, has had a profound impact on global health and economies. Understanding the trends in case numbers, testing rates, and positivity ratios is crucial for assessing the response effectiveness and identifying patterns in virus transmission. In this guided project, we analyze a dataset collected between January 20 and June 1, 2020, to investigate the relationship between the number of reported COVID-19 cases and the number of tests conducted across different countries.

Our objective is to determine which countries reported the highest number of positive cases relative to the number of tests performed. By leveraging data structures such as vectors, matrices, lists, and dataframes, we will conduct exploratory data analysis and derive meaningful insights from the dataset. This project serves as an opportunity to apply our data analysis skills in a real-world context and improve our understanding of the COVID-19 trends.

```{r}
# Load necessary libraries
library(readr)   # For reading CSV files
library(tibble)  # For the glimpse() function
library(dplyr)
library(ggplot2)
```


```{r}
# Load the dataset using the full path
covid_df <- read_csv("/home/student24/Documents/data science/Personal Projects/Personal researches/Coop Datasets/Covid_19/covid19.csv")

# Determine the dimensions of the dataset
dim(covid_df)

```

```{r}
# Get the column names
vector_cols <- colnames(covid_df)
print(vector_cols)  # Display column names
# Check the data structure of vector_cols
class(vector_cols)  # Should return "character" because it's a vector

# Display the first few rows of the dataset
head(covid_df)

# Display dataset summary
glimpse(covid_df)

```
```{r}
colnames(covid_df)

```

```{r}
# Filter only country-level data (removing state/province level data)
covid_df_all_states <- covid_df %>%
  filter(Province_State == "All States") %>% 
  select(-Province_State)  # Drop the Province_State column

# Display first few rows
head(covid_df_all_states)

# Check structure of the cleaned dataset
glimpse(covid_df_all_states)
```
In this step, we are focusing on working with daily data for our analysis. The dataset contains both cumulative data (e.g., total positive cases, total deaths) and daily data (e.g., daily positive cases, daily tests conducted). To avoid mixing these two types of data, we want to select only the columns related to daily measures.

The columns related to daily measures that we want to keep include:

- Date: The date of the observation.

- Country_Region: The country name.

- active: The number of active cases reported on that day.

- hospitalizedCurr: The number of actively hospitalized cases reported on that day.

- daily_tested: The number of tests conducted on that specific day.

-daily_positive: The number of new positive cases reported on that specific day.

We will store the result in a new dataframe called covid_df_all_states_daily.
```{r}
# Select the daily measures columns from covid_df_all_states
covid_df_all_states_daily <- covid_df_all_states %>%
  select(Date, Country_Region, active, hospitalizedCurr, daily_tested, daily_positive)

# Display the first few rows of the selected dataset to confirm the changes
head(covid_df_all_states_daily)

# Check the structure of the dataset to ensure only the intended columns are selected
glimpse(covid_df_all_states_daily)

```


```{r}
library(dplyr)


# Randomly select 5 countries
selected_countries <- covid_df_all_states_daily %>%
  distinct(Country_Region) %>%
  sample_n(5) %>%
  pull(Country_Region)

# Filter dataset for selected countries
covid_selected <- covid_df_all_states_daily %>%
  filter(Country_Region %in% selected_countries)

# Plot daily positive cases for selected countries
ggplot(covid_selected, aes(x = Date, y = daily_positive, color = Country_Region, group = Country_Region)) +
  geom_line(size = 1) +
  labs(title = "Daily Positive COVID-19 Cases for 5 Randomly Selected Countries",
       x = "Date",
       y = "Daily Positive Cases",
       color = "Country") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") + # Format dates for readability
  scale_y_continuous(labels = scales::comma) # Format y-axis with commas

```
##Interpretation: 
From the randomly selected countries above we notice an increase in positive cases from april to may.


In this step, we aim to compute the total number of tested, positive, active, and hospitalized cases by country, using the covid_df_all_states_daily dataframe, which contains daily data. We will then extract the top 10 countries based on the total number of tests conducted.

We need to:

1- Group the data by the Country_Region column, as we want to calculate the totals for each country.

2- Summarize the data by computing the sum for each of the relevant columns (tested, positive, active, hospitalized).

3-  Arrange the resulting data in descending order of the number of tests conducted.

4- Extract the top 10 countries based on the number of tests.


```{r}
# Summarize the data by computing the total for each country
covid_df_all_states_daily_sum <- covid_df_all_states_daily %>%
  group_by(Country_Region) %>%
  summarize(
    tested = sum(daily_tested, na.rm = TRUE),       # Total number of tests
    positive = sum(daily_positive, na.rm = TRUE),   # Total number of positive cases
    active = sum(active, na.rm = TRUE),             # Total number of active cases
    hospitalized = sum(hospitalizedCurr, na.rm = TRUE)  # Total number of hospitalized cases
  ) %>%
  # Arrange the data by the total number of tests (in descending order)
  arrange(desc(tested))


# Display the summarized data
print(covid_df_all_states_daily_sum)

# Extract the top 10 countries with the most tests
covid_top_10 <- head(covid_df_all_states_daily_sum, 10)

# Display the top 10 countries
print(covid_top_10)

```


```{r}

# Create a bar chart of tested vs positive cases
ggplot(covid_top_10, aes(x = reorder(Country_Region, -tested), y = tested, fill = "Tests")) +
  geom_bar(stat = "identity") +
  geom_bar(aes(y = positive, fill = "Positive Cases"), stat = "identity") +
  labs(
    title = "COVID-19 Tests vs. Positive Cases (Top 10 Countries)",
    x = "Country",
    y = "Number of Cases"
  ) +
  theme_minimal() +
  scale_fill_manual(name = "Case Type", values = c("Tests" = "blue", "Positive Cases" = "red")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
###Interpretation:
The number of postive cases in united state is higher as compard to those of other countries. India in this period had no positive case but has sever people going for testing


```{r}
# Pie chart for positive cases
ggplot(covid_top_10, aes(x = "", y = positive, fill = Country_Region)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(
    title = "Proportion of Positive Cases by Country",
    fill = "Country"
  ) +
  theme_void()

```

In order to store all relevant information for the top three countries with the highest positive-to-tested ratios in a structured format. The best way to do this is by creating a matrix that contains:

1-The positive/tested ratio

2-The total number of tested cases

3-The total number of positive cases

4-The total number of active cases

5-The total number of hospitalized cases


```{r}

# Creating vectors for each country
united_kingdom <- c(0.11, 1473672, 166909, 0, 0)
united_states <- c(0.10, 17282363, 1877179, 0, 0)
turkey <- c(0.08, 2031192, 163941, 2980960, 0)

# Creating a matrix combining the vectors
covid_mat <- rbind(united_kingdom, united_states, turkey)

# Renaming columns for better readability
colnames(covid_mat) <- c("Ratio", "tested", "positive", "active", "hospitalized")

# Displaying the matrix
print(covid_mat)


```

Combining all the results for easy retrieval involves

```{r}
countries <- c("United Kingdom", "United States", "Turkey")
tested_cases <- c(1473672, 17282363, 2031192)
positive_cases <- c(166909, 1877179, 163941)

# Calculate the ratio of positive cases to tested cases
positive_ratios <- positive_cases / tested_cases

# Store the top 3 as a named vector
positive_tested_top_3 <- setNames(positive_ratios, countries)

```


```{r}
# Step 1: Store the question
question <- "Which countries have had the highest number of positive cases against the number of tests?"

# Step 2: Store the answer
answer <- list("Positive tested cases" = positive_tested_top_3)

# Step 3: Create lists of data structures
dataframes_list <- list(covid_df, covid_df_all_states, covid_df_all_states_daily, covid_top_10)
matrices_list <- list(covid_mat)
vectors_list <- list(vector_cols, countries)

# Step 4: Combine all structures into a single list
data_structure_list <- list("Dataframes" = dataframes_list, 
                            "Matrices" = matrices_list, 
                            "Vectors" = vectors_list)

# Step 5: Create the final analysis list
covid_analysis_list <- list("Question" = question, 
                            "Answer" = answer, 
                            "Data Structures" = data_structure_list)

# Step 6: Display the answer
print(covid_analysis_list[["Answer"]])


```
## Interpretation of Positive Tested Cases Ratio

The following analysis examines the proportion of positive COVID-19 cases relative to the total number of tested cases in three countries: **United Kingdom, United States, and Turkey**. 

- **United Kingdom** has the highest positive test ratio of **11.33%**, meaning that for every 100 individuals tested, approximately **11** were positive.
- **United States** follows closely with a ratio of **10.86%**, indicating that around **11 out of 100** tested individuals were positive.
- **Turkey** has a lower ratio of **8.07%**, meaning that about **8 out of 100** tested individuals were positive.

### **Implications:**
- A **higher ratio** may suggest that testing was focused on symptomatic individuals, or that the virus was more widespread in that country.
- A **lower ratio** could indicate broader testing, including asymptomatic individuals, or better outbreak control.

## Bar Plot of Positive Tested Cases Ratio

```{r positive_tested_plot, echo=TRUE, message=FALSE, warning=FALSE}
# Create a data frame for visualization
positive_tested_df <- data.frame(
  Country = c("United Kingdom", "United States", "Turkey"),
  Positive_Tested_Ratio = c(0.1133, 0.1086, 0.0807) # Rounded values
)

# Create a bar plot
ggplot(positive_tested_df, aes(x = Country, y = Positive_Tested_Ratio, fill = Country)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportion of Positive COVID-19 Cases vs. Tested Cases",
    x = "Country",
    y = "Positive Tested Ratio"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red", "orange"))
```

